Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-27 ms/spaceness.tex
edited until discussion

\ak{I think we can cut the following paragraph if we just add the last sentence above} 
\cjb{agree - commenting out}
%Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, the probability of survival until the next time step for individual $i$ is
\begin{equation}
    p_i \min(0.95, \frac{1}{1 + n_i / (K (1+L)} .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is around $K$,
and the mean lifetime is around $L$.

@@

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e. most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.

@@

\subsection{Genealogical Parameters}
Having implemented our continuous space forward time population genetic simulation, the first thing we were interested in examining was the effect of spatial parameters on population parameters such as the generation time, the census population size, and the variance in offspring number. In contrast to intuition based on standard population genetic models, in the context of our non-Wright-Fisher model basic quantities like the generation time depend on life history parameters such as the dispersal distance and carrying capacity of the landscape. We first examined how the population size, generation time, and  variance in number of offspring vary with neighborhood size, which itself is a function of $\sigma$ (Figure \ref{fig:genparams}). There are a few things to note. First all three quantities are non-linear with respect to neighborhood size. Census size largely declines as neighborhood size increases for both the spatial and random mating models. However for spatial models this decline only begins for neighborhood size $\geq 10$. By a neighborhood size $\geq 100$ the spatial and random mating models are indistinguishable from one another, a sign that our simulations are performing as expected. Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.

@@

Generation times similarly show complex behavior with respect to neighborhood sizes, and vary across the parameter range explored between 5.2 and 4.9 timesteps per generation. To stress this point further, the generation time here varies because under our non-Wright-Fisher dynamics individuals are free to reproduce more than once over their lifespan and the lifespan of an individual is only indirectly determined by the parameterization of density-dependent competition the model. Under both the spatial and random mating regimes generation time reaches a nadir at a neighborhood size of $\approx 50$. Interestingly under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.

Last we looked at the effect of neighborhood size on the variance in number of offspring in the population -- a key parameter determining the effective population size. Surprisingly the spatial and random mating model behave quite differently with respect to the neighborhood size: while the variance in offspring number increases monotonically, or nearly so, under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.

@@

\subsection{Impacts of Continuous Space on Population Genetic Summary Statistics}
One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation. Indeed we have little knowledge to date about how some of our most beloved population genetic summary statistics might be affected by such realistic models. Moreover, as we will show, sampling strategies of individuals with respect to space can affect summaries of variation at least as strongly as the underlying population dynamics.

@@

\subsubsection{Site Frequency Spectra and Summaries of Diversity}
In Figure \ref{fig:sumstats} we examine the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure \ref{fig:sumstats}A) and several canonical population genetic summary statistics (Figure \ref{fig:sumstats}B). For populations in continuous space we observe a significant skew in the SFS for smaller neighborhood sizes ($\leq 100$) that is exacerbated by mid point and point sampling of individuals (see Figure \ref{fig:samplemap}). As expected these effects are strongest for the smallest neighborhood sizes examined and are consistently in the direction of an enrichment of intermediate frequency variants in comparison to the standard neutral expectation. This can be seen succinctly in the response of Tajima's \textit{D} to variation in neighborhood sizes and sampling regime shown in Figure \ref{fig:sumstats}B, whereby we see \textit{D} turning quite positive at small neighborhood sizes. This effect is particularly strong for the midpoint and point sampling regimes, but also occurs under random sampling. Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies -- varying from a neighborhood size of $\approx50$ for random sampling to $\geq1000$ for midpoint sampling.

One of the most commonly used summaries of variation is Tajima's estimator of the neutral mutation rate $\widehat{\theta}_{\pi}$. As can be seen in Figure \ref{fig:sumstats}B $\widehat{\theta}_{\pi}$ can vary as much as nearly three-fold for the smallest neighborhood sizes between the random mating and spatial models depending on the sampling strategies. Each sampling strategy approaches the random mating expectation at its own rate, but by neighborhood size of $\approx 100$ all models are equivalent. These results are to be expected as at small neighborhood sizes the time to the most recent common ancestor for the population should be pushed back-- i.e. it takes longer for lineages to coalesce if dispersal is limited. Interestingly the rate of approach to random mating expectations across sampling strategies is reversed relative to that observed in Tajima's D -- midpoint sampling reaches random mating expectations at neighborhood sizes $\approx50$ while random sampling is inflated until neighborhood size $/approx$ 100. In effect this means that any estimates of population size based on expected heterozygosity from populations with limited dispersal should be treated with caution.

Patterns from observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. $F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. Indeed for random sampling under the spatial model $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. The patterns for raw observed heterozygosity are more complex. Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, as a result of the a higher proportion of heterozygotesn occurring in the middle of the landscape (Figure S\ref{fig:hetmap}). This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

Trends in pairwise haplotype block sharing parallel those in allele-frequency-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}). At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values with respect to the random mating expectation-- resulting in lower means and fewer long IBS blocks. The variance and skew of the distribution of haplotype block lengths are only minorly affected by neighborhood size in our simulations when calculated across all pairs of individuals; however, they are strongly dependent on sampling regimes. For example, the number of long haplotype blocks declines as neighborhood size increases under midpoint sampling but changes very little across neighborhood sizes under point or random sampling. Thus sampling strategies with respect to geography will affect conclusions drawn from haplotype length distributions quite dramatically.

\subsubsection{Correlations of summary statistics with geographic distance}
Correlating population genetic summaries such as $F_{ST}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}. As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance among samples with a number of our summary statistics (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}). Starting with a measure of population differentiation,$D_{xy}$, we observe a positive correlation with distance that declines as dispersal increases, as expected under the theory developed by \citep{Rousset1997} and others. This relationship varies across sampling strategies of course, with the weakest correlations observed for midpoint sampling. Moreover there is clearly strong signal for  neighborhood size in the strength of the correlation between $D_{xy}$ and geographic distance.

We next turn our attention to the effect of geographic distance on haplotype block length sharing. As in \citep{Ringbauer2017} and \citep{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal. This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}). Of these summaries the mean of the IBS tract length distribution is only weakly affected by neighborhood size, likely because it is heavily influenced by the small number of very long IBS tracts. In contrast the count of long IBS blocks and the skew of the pairwise IBS block distribution are strongly dependent on distance among individuals, and the magnitude of this correlation declines predictably with neighborhood size. In all spatial correlations random and point sampling are similarly correlated with space across neighborhood sizes, but midpoint sampling causes weaker correlations because it incorporates less genetic and geographic distance than the full sample.

\subsubsection{Effects of Sampling on Random Mating and Spatial Models}
To summarize the effect of sampling strategy on genetic diversity in random mating and spatial models we also asked if summary statistics varied significantly across sampling regimes. In table \ref{table:sampling} we show that for spatial models the mean and variance of nearly all summary statistics was significantly different across sampling strategies (Table \ref{table:sampling}) for spatial models, but not for random mating models. In summary, sampling strategy can be safely ignored when the population is randomly mating but will shape estimates of genetic diversity in any population with limited dispersal.

@@

In Figure \ref{fig:demography} we show inferred population size histories from our simulations as a function of neighborhood size binned in to one of four categories. In general we observe that demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}). This is consistent with our observations of the SFS from which Stairwayplot is doing its inference. Inflated past population sizes were seen in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed among sampling regimes is enough to alter demographic estimates. More alarmingly, inference of severe population bottlenecks was  common at neighborhood sizes under 100 for midpoint and point sampling strategies. Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$. While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.  
To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation. Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process. Instead we found that standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}). In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot. This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.

@@

\subsection{GWAS}
To ask what confounding effects spatial genetic variation might have on genome-wide association studies we performed GWAS on our simulations using phenotypes that were determined solely by the environment. In general we found that for simulations with limited dispersal, i.e. neighborhood size $< 100$, spatial variation in the environment causes GWAS to infer significant associations with purely environmental phenotypes at over 25\% of sampled SNPs if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}). This effect is particularly strong for clinal and corner environments, for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations. Patchy environmental distributions, which result in fewer extreme phenotype values in the sample (Figure \ref{fig:gwas}A), cause fewer false-positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes. Notably no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.

When PCA positions are included as covariates to control for population structure in GWAS the vast majority of SNPs no longer surpass a 5\% FDR significance threshold, but up to 1.5\% of SNPs are significantly associated at low dispersal distances under corner and patchy environmental distributions (Figure \ref{fig:gwas}C). At neighborhood sizes $> 500$ up to 0.31\% of SNPs were significant for corner and clinal environments. Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.

Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure \ref{fig:gwas}. This appears to reflect a loss of descriptive power in the PCA -- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 0.1 to 0.04 (Figure \ref{fig:gwas}B). Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment. A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.

In Figure \ref{fig:gwas}D we present quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions. For clinal environments $-log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression. In contrast corner and patchy environments cause the greatest inflation in $-log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal. Finally, we observed that PC regression appears to cause some degree of overcorrection for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
\caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-log10(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
%A: example environmental distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: Proportion of significant SNPs after FDR correction for GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect deflated $p$ values.
}
\label{fig:gwas}
\end{figure*}

>>>>

\ak{I think we can cut the following paragraph if we just add the last sentence above} 
\cjb{agree - commenting out}
%Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e., the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, the probability of survival until the next time step for individual $i$ is
\begin{equation}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is around $K$,
and the mean lifetime is around $L$.

@@

We then estimated correlation coefficients between spatial distance and each moment of the pairwise IBS tract distribution. Because more closely related individuals on average share longer haplotype blocks we expect that spatial distance will be negatively correlated with mean haplotype block length, and that this correlation will be strongest (i.e., most negative) when dispersal is low. The variance, skew, and count of long haplotype block statistics are meant to reflect the relative length of the right (upper) tail of the distribution, which represents the frequency of long haplotype blocks so should reflect recent demographic events \citep{chapman2002effect}.

@@

%%%%%%%%%%%
\subsection{Demographic Parameters}

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number.
These are shown in Figure~\ref{fig:genparams}.
This occurs because,
even through the ``population density'' ($K$) and ``mean lifetime'' ($L$) parameters
were the same in all simulations,
the strength of stochastic effects depends strongly on $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
By a neighborhood sizes larger than 100, the spatial and random mating models are indistinguishable from one another, 
a sign that our simulations are performing as expected. 
Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000.

@@

Generation time similarly shows complex behavior with respect to neighborhood sizes, and varies between 5.2 and 4.9 timesteps per genration across the parameter range explored. 
% This occurs because individuals may reproduce more than once over their lifespan, and the lifespan of an individual is determined by density-dependent competition the model. 
Under both the spatial and random mating models, 
generation time reaches a minimum at a neighborhood size of around 50. 
Interestingly, under the range of neighborhood sizes that we examined, generation times between the random mating and spatial models are never quite equivalent -- presumably this would cease to be the case at neighborhood sizes higher than we simulated here.

Last, we looked at the variance in number of offspring -- a key parameter determining the effective population size. 
Surprisingly, the spatial and random mating models behave quite differently: 
while the variance in offspring number increases nearly monotonically under the spatial model, the random mating model actually shows a decline in the variance in offspring number until a neighborhood size $\approx 10$ before it increases and eventually equals what we observe in the spatial case.

@@

%%%%%%%%%%%%
\subsection{Impacts of Continuous Space on Population Genetic Summary Statistics}

Even though certain aspects of population demography depend on the scale of spatial interactions,
it still could be that population genetic variation is well-described by a well-mixed population model.
Indeed, mathematical results suggest that genetic variation in some spatial models should be well-approximated by a Wright-Fisher population
if neighborhood size is large and all samples are geographically widely separated \citep{wilkins2004separation,zahle2005stepping}.
% One of the main goals of this study is to examine the effect of continuous space on canonical summaries of population genetic variation. 
However, the behavior of most common population genetic summary statistics has not yet been described in realistic geographic models.
Moreover, as we will show, spatial sampling strategies can affect summaries of variation at least as strongly as the underlying population dynamics.

@@

%%%%%%%%%%%%%%%%
\subsubsection{Site Frequency Spectra and Summaries of Diversity}

Figure \ref{fig:sumstats} shows the effect of varying neighborhood size and sampling strategy on the site frequency spectrum (Figure~\ref{fig:sumstats}A) 
and several standard population genetic summary statistics (Figure~\ref{fig:sumstats}B). 
These show a significant enrichment of intermediate frequency variants in comparison to the nonspatial expectation,
for smaller neighborhood sizes ($\leq 100$) that is exacerbated by midpoint and point sampling of individuals (depicted in Figure~\ref{fig:samplemap}). 
Reflecting this, Tajima's $D$ is quite positive in the same situations (Figure~\ref{fig:sumstats}B).
Notably, the point at which Tajima's $D$ approaches 0 differs strongly across sampling strategies 
-- varying from a neighborhood size of roughly 50 for random sampling to at least 1000 for midpoint sampling.

One of the most commonly used summaries of variation is Tajima's summary of nucleotide divergence, $\widehat{\theta}_{\pi}$,
\plr{In the figure this has no hat on; should we remove these in the text? The notation is not standard.}
calculated as the mean density of nucleotide differences, averaged across pairs of samples.
As can be seen in Figure \ref{fig:sumstats}B, $\widehat{\theta}_{\pi}$ differs as much as nearly three-fold 
between the random mating and spatial models.
$\widehat{\theta}_{\pi}$ using each sampling strategy approaches the random mating expectation at its own rate, 
but by a neighborhood size of around 100 all models are equivalent. 
\plr{To interpret this we need to compare to census size over variance in offspring number.}
The differences between spatial and random mating simulations are much greater than expected 
from differences in census size divided by variance in offspring number.
This likely occurs because $\widehat{\theta}_{\pi}$ is a measure of mean time to most recent common ancestor between two samples,
and at small values of $\sigma$, the time for dispersal to mix ancestry across the range exceeds the mean nonspatial coalescent time.
(For instance, at the smallest value of $\sigma=0.2$, the range is 250 dispersal distances wide,
and since the location of a diffusively moving lineage after $k$ generations has variance $k \sigma^2$, 
it takes around $250^2 = 62500$ generations to mix across the range,
which is roughly ten times larger than the random mating effective population size.)
Interestingly, the effect of sampling strategy is reversed relative to that observed in Tajima's D 
-- midpoint sampling reaches random mating expectations around neighborhood size 50, 
while random sampling is inflated until around neighborhood size 100. 
% In effect, this means that any estimates of population size based on expected heterozygosity from populations with limited dispersal should be treated with caution.

Values of observed heterozygosity and its derivative $F_{IS}$ also depend heavily on neighborhood size under spatial models as well as the sampling scheme. 
$F_{IS}$ is inflated above the expectation across most of the parameter space examined and across all sampling strategies. 
This effect is caused by a deficit of heterozygous individuals in low-dispersal simulations 
-- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. 
Indeed, for random sampling under the spatial model, $F_{IS}$ does not approach the random mating equivalent until neighborhood sizes of nearly 1000. 
On the other hand, the dependency of raw observed heterozygosity on neighborhood size is not monotone.
Under midpoint sampling observed heterozygosity is inflated even over the random mating expectation, 
as a result of the a higher proportion of heterozygotes occurring in the middle of the landscape (Figure~\ref{fig:hetmap}). 
This squares with a report from \cite{Shirk2014} who observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

\plr{peter to revisit this paragraph}
Trends in pairwise haplotype block sharing parallel those in allele-frequency-based diversity estimates (Figure \ref{fig:sumstats}, Supplementary Figure \ref{fig:allsumstats}). At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values with respect to the random mating expectation -- resulting in lower means and fewer long IBS blocks. The variance and skew of the distribution of haplotype block lengths are only minorly affected by neighborhood size in our simulations when calculated across all pairs of individuals; however, they are strongly dependent on sampling regimes. For example, the number of long haplotype blocks declines as neighborhood size increases under midpoint sampling but changes very little across neighborhood sizes under point or random sampling. Thus sampling strategies with respect to geography will affect conclusions drawn from haplotype length distributions quite dramatically.

\subsubsection{Correlations of summary statistics with geographic distance}
Correlating population genetic summaries such as $F_{ST}$ against geographic distance has shown great utility in empirical population genetics \citep{Rousset1997}. As we know the exact locations of individuals that are sampled in our simulated populations we have examined the relationship of geographic distance between samples with a number of summary statistics (Figures \ref{fig:sumstats} and \ref{fig:allsumstats}). 
Mean density of nucleotide differences between individuals, $D_{xy}$, is positively correlated with the geographic distance between the individuals,
and the strength of this correlation declines as dispersal increases, as expected under theory \citep{Wright1943,Rousset1997}. 
This relationship varies across sampling strategies, with the weakest correlations observed for midpoint sampling,
perhaps due to a dearth of long-distance comparisons.
% Moreover there is clearly strong signal for neighborhood size in the strength of the correlation between $D_{xy}$ and geographic distance.

\plr{peter to revisit this paragraph}
We next turn our attention to the effect of geographic distance on haplotype block length sharing. 
As in \citet{Ringbauer2017} and \citet{Baharian2016} we found that the pairwise distribution of haplotype block lengths is more strongly left-skewed under limited dispersal. This is reflected in negative correlation coefficients between spatial distance and the mean, variance, skew, and count of long blocks from the pairwise distribution of identical-by-state block lengths (Figure \ref{fig:sumstats} and Figure \ref{fig:allsumstats}). Of these summaries the mean of the IBS tract length distribution is only weakly affected by neighborhood size, likely because it is heavily influenced by the small number of very long IBS tracts. In contrast the count of long IBS blocks and the skew of the pairwise IBS block distribution are strongly dependent on distance among individuals, and the magnitude of this correlation declines predictably with neighborhood size. In all spatial correlations random and point sampling are similarly correlated with space across neighborhood sizes, but midpoint sampling causes weaker correlations because it incorporates less genetic and geographic distance than the full sample.

%%%%%%%%%%%
\subsubsection{Spatial distribution of allele copies}
\plr{insert bit describing that plot here}

@@

\plr{revsit with smc++}
Figure \ref{fig:demography} shows inferred population size histories, grouped by neighborhood size and sampling strategy.
In general, Stairwayplot tends to infer ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure~\ref{fig:demography}). 
This is consistent with our observations of the SFS from which Stairwayplot is doing its inference. Inflated past population sizes were seen in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed among sampling regimes is enough to alter demographic estimates. More alarmingly, inference of severe population bottlenecks was  common at neighborhood sizes under 100 for midpoint and point sampling strategies. Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$. While that is so individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.  
We compared this range of variation to Stairwayplot results run on coalescent simulations with constant population size,
and found that the noisiest results ($N_W < 20$, or midpoint sampling and $N_W < 100$)
were noiser than expected, but remaining simulation results were consistent with Stairwayplot's behavior on data from random mating models (Table~\ref{table:demography}).
% To test whether the variation we observed in inferred demographic histories from Stairwayplot was the result of spatial effects in our simulations rather than the behavior of the optimization routine we compared standard deviations of inferred population sizes in each sampling/neighborhood size bin with those returned by an equilibrium coalescent simulation. Were these to be equal we could assume that the variation were purely as to be expected under the noise inherent in the genealogical process. Instead we found that standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model (Table \ref{table:demography}). 
In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot. This underscores the fact that some \textit{a priori} knowledge about the population dynamics at play will be important to interpreting results of demographic estimation routines.

@@

%%%%%%%%%%%%%%%%%
\subsection{GWAS}

To ask what confounding effects spatial genetic variation might have on genome-wide association studies we performed GWAS on our simulations using phenotypes that were determined solely by the environment
-- so, any SNP showing statistically significant correlation with phenotype is a false positive. 
As expected, spatial autocorrelation in the environment causes spurious associations 
across much of the genome if no correction for genetic relatedness among samples is performed (Figure \ref{fig:gwas}). 
This effect is particularly strong for clinal and corner environments, 
for which the lowest dispersal levels cause over 60\% of SNPs in the sample to return significant associations. 
Patchy environmental distributions, which is less strongly spatially correlated (Figure \ref{fig:gwas}A), 
cause fewer false positives overall but still produce spurious associations at roughly 10\% of sites at the lowest neighborhood sizes. 
Notably, no simulations with nonspatial environments returned more than one significant association (Figure \ref{fig:gwas}C), 
demonstrating that this effect is caused specifically by the interaction of population structure and spatial variation in the environment rather than by population structure itself.

The confounding effects of geographic structure are well-known, 
and it is common practice to control for this by including principal components (PCs) as covariates to control for these effects.
This mostly works in our simulations -- after doing this, the vast majority of SNPs no longer surpass a 5\% FDR significance threshold. 
However, a substantial number of SNPs -- up to 1.5\% of SNPs -- still surpass this threshold (and thus would be false positives in a GWAS),
especially under ``corner'' and ``patchy'' environmental distributions (Figure~\ref{fig:gwas}C). 
At neighborhood sizes larger than 500, up to 0.31\% of SNPs were significant for corner and clinal environments. 
Given an average of 132,000 SNPs across simulations after MAF filtering, this translates to up to 382 false-positive associations. 
In most cases the $p$ values for these associations were significant after FDR correction but would not pass the threshold for significance under the more conservative Bonferroni correction, suggesting that the most strongly associated variants from many GWAS of mono- or oligogenic traits are robust to this variety of stratification bias.

Clinal environments cause an interesting pattern in false positives after PC correction: at low neighborhood sizes the correction removes nearly all significant associations, 
but at neighborhood sizes above $\approx250$ the proportion of significant SNPs increases to up to 0.4\% (Figure~\ref{fig:gwas}). 
This may be due to a loss of descriptive power of the PCs
-- as neighborhood size increases, the total proportion of variance explained by the first 10 PC axes declines from roughly 10\% to 4\% (Figure \ref{fig:gwas}B). 
Essentially, PCA seems unable to effectively summarize the weak population structure present in large-neighborhood simulations, but these populations continue to have enough spatial structure to create significant correlations between genotypes and the environment. A similar process can also be seen in the corner phenotype distribution, in which the count of significant SNPs initially declines as neighborhood size increases and then increases at approximately the point at which the proportion of variance explained by PCA approaches its minimum.

Figure \ref{fig:gwas}D shows quantile-quantile plots that show the degree of genome-wide inflation of test statistics in PC-corrected GWAS across all simulations and environmental distributions. 
For clinal environments, $-\log_{10}(p)$ values are most inflated when neighborhood sizes are large, consistent with the pattern observed in the count of significant associations after PC regression. In contrast corner and patchy environments cause the greatest inflation in $-\log_{10}(p)$ at neighborhood sizes $<100$, which likely reflects the inability of PCA to account for fine-scale structure caused by very limited dispersal. Finally, we observed that PC regression appears to overfit to some degree for all phenotype distributions, visible in Figure \ref{fig:gwas}D as points falling below the 1:1 line.

\plr{Add a few manhattan plots to the supplement and refer to them somewhere.}

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary.pdf}
\caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding PC positions as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-log10(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
%A: example environmental distributions and sampling maps, with colors scaled to phenotype values. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: Proportion of significant SNPs after FDR correction for GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed $p$ values relative to those expected from a uniform distribution. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect deflated $p$ values.
    \plr{suggestion: (a) log-scale the y axes in (C) and add another y-axis on the right showing (rough number of human genome SNPs above 5\% times this percentage) so we can see absolute numbers}
}
\label{fig:gwas}
\end{figure*}
